Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-15 submission/manuscript.Rmd
fix formatting on fig6 caption

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis.]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} \textbf{A.}  ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom). \textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS $\pm$ 500) in EpiLCs (purple) and PNCs (blue). \textbf{C.} Gene ontology (GO) comparision of genes with KDM5C bound to their promoter in EpiLCs and PNCs. Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared). \textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right). \textbf{E.} Eulerr overlap of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple). \textbf{F.} Example KDM5C ChIP-seq signal around the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs. \textbf{G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs. \textbf{H.} Number of all gene promoters bound or unbound by KDM5C with instances of the E2F or E-box consensus sequence. \textbf{I.} HOMER motif analysis of all KDM5C-unbound germline gene promoters, highlighting significant enrichment of RFX family transcription factor motifs. \textbf{J.} Expression of RNA-seq DEG \textit{Rfx2} in wild-type and \textit{Kdm5c}-KO EpiLCs. P-value of Welch's t-test, expression in transcripts per million (TPM).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} \textbf{A.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation and collection time points. \textbf{B.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs at 0, 24, and 48 hours of differentiation into EpiLCs.Expression calculated in comparision to TBP mRNA expression (2\textsuperscript{-deltaCT}). * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{C.} KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - representative lanes of Western blot for KDM5C and DAXX. * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{D.} Top - Representative UCSC browser view of histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq signal at two germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs. Bottom - Average H3K4me2 signal at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs. \textbf{E.} Top - Representative UCSC browser view of histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq signal at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala. Bottom - Average H3K4me3 signal at the transcripiton start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala. \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters (CGI promoter, red), based on UCSC annotation. Left - percentage of all germline-enriched genes, middle - KDM5C-bound germline genes, and right - KDM5C-unbound germline genes \textbf{G.} enrichPlot gene ontology analysis of germline genes with (CGI-promoter) or without (CGI-free) CGIs in their promoter. \textbf{H.} Expression of germline genes with (CGI-promoter, red) or without (CGI-free, salmon) CGIs in their promoter across stages of spermatogenesis from Green et al 2018. SPG - spermatogonia, Pre-Lep - preleptotene spermatocytes, S-Cytes - meiotic spermatocytes, S-Tids - post-meiotic haploid round spermatids, E.Sp - elongating spermatids. Wilcoxon test, * p < 0.05, ** p < 0.01, *** p < 0.001, **** p < 0.0001. \textbf{I.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation \textbf{J.} Volcano plot of whole genome bisulfite sequencing (WGBS) comapring CpG methylation (CpGme) at germline gene promoters (TSS $\pm$ 500) in wild-type (WT) ESCs versus exEpiLCs. Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25%) with CGIs in red, CGI-free promoters in light blue \textbf{K.} Volcano plot of WGBS of wild-type (WT) versus \textit{Kdm5c}-KO exEpiLCs for germline gene promoters. Promoters with CGIs in red, CGI-free promoters in light blue. \textbf{L.} Histogram of average percent CpGme at the promoter of germline genes with or without CGIs. Wild-type in navy and \textit{Kdm5c}-KO (KO) in light blue. Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test. \textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs}
  \label{figurelabel}
\end{figure}

>>>>

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/KDM5C_ChIPseq.pdf}
  \caption[KDM5C binds to a subset of germline gene promoters during early embryogenesis.]{\textbf{KDM5C binds to a subset of germline gene promoters during early embryogenesis.} \textbf{A.}  ChIPseeker localization of KDM5C peaks at different genomic regions in EpiLCs (top) and hippocampal and cortex primary neuron cultures (PNCs, bottom). \textbf{B.} Overlap of genes with KDM5C bound to their promoters (TSS $\pm$ 500) in EpiLCs (purple) and PNCs (blue). \textbf{C.} Gene ontology (GO) comparison of genes with KDM5C bound to their promoter in EpiLCs and PNCs. Genes were classified as either bound in EpiLCs only (EpiLC only), unique to PNCs (PNC only, no significant ontologies) or bound in both PNCs and EpiLCs (shared). \textbf{D.} Average KDM5C binding around the transcription start site (TSS) of all germline-enriched genes in EpiLCs (left) and PNCs (right). \textbf{E.} Eulerr overlap of germline-enriched genes (green) with significant KDM5C binding at their promoter in EpiLCs (purple). \textbf{F.} Example KDM5C ChIP-seq signal around the \textit{Dazl} TSS but not \textit{Stra8} in EpiLCs. \textbf{G.} HOMER motif analysis of all KDM5C-bound germline gene promoters, highlighting significant enrichment of MAX, E2F6, and E2F motifs. \textbf{H.} Number of all gene promoters bound or unbound by KDM5C with instances of the E2F or E-box consensus sequence. \textbf{I.} HOMER motif analysis of all KDM5C-unbound germline gene promoters, highlighting significant enrichment of RFX family transcription factor motifs. \textbf{J.} Expression of RNA-seq DEG \textit{Rfx2} in wild-type and \textit{Kdm5c}-KO EpiLCs. P-value of Welch's t-test, expression in transcripts per million (TPM).}
  \label{figurelabel}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{../submission/compiled_figs/KDM5C_Mechanism.pdf}
  \caption[KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} \textbf{A.} Diagram of embryonic stem cell (ESC) to epiblast-like cell (EpiLC) differentiation and collection time points. \textbf{B.} Real time quantitative PCR (RT-qPCR) of \textit{Kdm5c} mRNA expression in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs at 0, 24, and 48 hours of differentiation into EpiLCs. Expression calculated in comparison to TBP mRNA expression (2\textsuperscript{-deltaCT}). * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{C.} KDM5C protein expression normalized to DAXX. Quantified intensity using ImageJ (artificial units - au). Right - representative lanes of Western blot for KDM5C and DAXX. * p < 0.05, ** p < 0.01, *** p < 0.001, Welch's t-test. \textbf{D.} Top - Representative UCSC browser view of histone 3 lysine 4 dimethylation (H3K4me2) ChIP-seq signal at two germline genes in wild-type and \textit{Kdm5c}-KO EpiLCs. Bottom - Average H3K4me2 signal at the TSS of all germline-enriched genes in wild-type (dark purple) and \textit{Kdm5c}-KO (light purple) EpiLCs. \textbf{E.} Top - Representative UCSC browser view of histone 3 lysine 4 trimethylation (H3K4me3) ChIP-seq signal at two germline genes in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) adult amygdala. Bottom - Average H3K4me3 signal at the transcription start site (TSS) of all germline-enriched genes in wild-type (dark red) and \textit{Kdm5c}-KO (light red) amygdala. \textbf{F.} Percentage of germline genes that harbor CpG islands (CGIs) in their promoters (CGI promoter, red), based on UCSC annotation. Left - percentage of all germline-enriched genes, middle - KDM5C-bound germline genes, and right - KDM5C-unbound germline genes. (Legend continued on next page.)}
  \label{figurelabel}
\end{figure}
\newpage
\addtocounter{figure}{-1}
\begin{figure} [t!]
  \caption[]{\textbf{KDM5C promotes long-term silencing of germline genes via DNA methylation at CpG islands.} (Legend continued.) \textbf{G.} enrichPlot gene ontology analysis of germline genes with (CGI-promoter) or without (CGI-free) CGIs in their promoter. \textbf{H.} Expression of germline genes with (CGI-promoter, red) or without (CGI-free, salmon) CGIs in their promoter across stages of spermatogenesis from Green et al 2018. SPG - spermatogonia, Pre-Lep - preleptotene spermatocytes, S-Cytes - meiotic spermatocytes, S-Tids - post-meiotic haploid round spermatids, E.Sp - elongating spermatids. Wilcoxon test, * p < 0.05, ** p < 0.01, *** p < 0.001, **** p < 0.0001. \textbf{I.} Diagram of ESC to extended EpiLC (exEpiLC) differentiation. \textbf{J.} Volcano plot of whole genome bisulfite sequencing (WGBS) comparing CpG methylation (CpGme) at germline gene promoters (TSS $\pm$ 500) in wild-type (WT) ESCs versus exEpiLCs. Significantly differentially methylated promoters (q < 0.01, |methylation difference| > 25\%) with CGIs in red, CGI-free promoters in light blue \textbf{K.} Volcano plot of WGBS of wild-type (WT) versus \textit{Kdm5c}-KO exEpiLCs for germline gene promoters. Promoters with CGIs in red, CGI-free promoters in light blue. \textbf{L.} Histogram of average percent CpGme at the promoter of germline genes with or without CGIs. Wild-type in navy and \textit{Kdm5c}-KO (KO) in light blue. Dashed lines are average methylation for each genotype, p-values for Mann-Whitney U test. \textbf{M.} UCSC browser view of around the transcription start site (TSS) of germline genes, showing UCSC annotated CGI, representative CpGme in wild-type (WT) and \textit{Kdm5c}-KO (5CKO) ESCs and exEpiLCs.}
\end{figure}
